Glyph-maps for Visualizing Climate Data and Models

### 2011-09-29 paper.tex
data refs, nz to us spelling

Climate data is composed of measurements, such as temperature, precipitation, and winds, with a spatial and temporal context. The classic display for data of this type of is ``heatmap'', a tiled plot using color to display the value of the variable of interest as each location. When measurements are made at multiple time points it is common to display the data with small multiples, such as in Figure~\ref{fig:nasa-facet}. In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America \citep{murrell:2010}. Color is used to display de-seasonalized temperature, with red mapped to high values and blue to low. The most noticeable feature is the strong red patch in the equatorial Pacific beginning mid-1997 and tapering out during 1998. This is the El Ni\~no event, a major temperature anomaly. More work is required to see more localized patterns, like the cooler land temperatures seen in early years.

@@

We begin in Section~\ref{sec:construction} by describing the algorithm used to create glyphs-maps. Section~\ref{sec:perception} discusses their perceptual properties, including the importance of a visual reference grid, and of carefully considering scale. Section~\ref{sec:large-data} shows how glyph-maps scale to large data, and the interplay of models with glyph-maps. We originally developed glyph-maps for data on a rectangular grid, but many spatiotemporal data sets have irregular spatial locations. Section~\ref{sec:irregular} discusses how glyph-maps can be adjusted for this type of data.

To demonstrate the glyph-map we will use three datasets:

  \item The ASA 2009 data expo data which consists of monthly observations of
  several atmospheric variables from the International Satellite Cloud
  Climatology Project. The dataset includes observations over 72 months
  (1995--2000) on a 24 x 24 grid (576 locations) stretching from
  $113.75^{\circ}$W to $56.25^{\circ}$W longitude and $21.25^{\circ}$S to
  $36.25{^\circ}$N latitude.

\item GISTEMP surface temperature data provided on $2^{\circ}$ x
  $2^{\circ}$ grid over the entire globe, measured monthly. Ground
  station data was de-seasonalized, differenced from from the
  1951-1980 temperature averages, and spatially averaged to obtain
  gridded measurements. For the purposes of this paper, we extracted
  the locations corresponding to the continental USA. The data is made
  available by the Earth System Research Laboratory, Physical Sciences
  Division, National Oceanic and Atmospheric Administration, from
  their web site at \url{http://www.esrl.noaa.gov/psd/}.

  \item USHCN (Version 2) ground station network. USHCN data provided by NOAA NCDC from their website \url{http://cdiac.ornl.gov/ftp/ushcn_v2_monthly/}.

@@

Since the coordinates are linear combinations of two variables, it is possible to build these these line-glyphs interactively, illustrated by software packages that have incorporated tours \citep{cook:2006}, such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}. This technique is used to good effect in \citet{buja:1996a}, and was used to examine the climate data in \citet{hobbs:2010}.

@@

An third alternative is to calculate the relevant statistic for each location and display these values as single heatmap. For example, to study long-term trends, we could calculate the slope of a linear model at each location and color tiles across the map by the value of the slope. The disadvantage of this approach is that it is not exploratory, and requires that we have a very precise definition of the measurement we are interested in, and it still suffers from the relative difficultly of colour perception.

@@

Each glyph is small, so there is not enough space for a full set of axes. Instead, we add reference lines and boxes. These need to be minimally perceptible, post-attentive \citep{healey} and de-emphasized, so as not to detract from the raw data. The \textbf{reference grid} is a box, framing each icon, which represents the spatial grid. The \textbf{reference line} is a horizontal line at mid-range. Both help to read differences in slope and intercept. Slope is read by comparing the position of left and right endpoints on the frame, or by reading the angle between the data line and the reference line. Intercept is read from ``average'' position of the line in the box: is it near the top, or near the bottom? Figure~\ref{fig:ref-basic} demonstrates both guides. We draw them in white, so they are minimally perceptible relative to the black data.

@@

%It's also often useful to display other types of reference information. You are only limited by your imagination (and the capabilities of your graphics software), but two simple ideas are to (1) vary the colour or transparency of glyphs or references, or (2) use an additional layer to highlight special points. Figure~\ref{fig:ref-adv} illustrates these ideas to display the month with the highest temperature at each locations. Careful colour choice is necessary so that this information is perceptible, but not distracting.

%\begin{figure}[htbp]
%  \centering
%  \includegraphics[width=0.5\linewidth]{ref-max-1}%
%  \includegraphics[width=0.5\linewidth]{ref-max-2}
%  \caption{Glyph-maps of seasonal temperature patterns also showing the month with the highest temperature using (left) an additional layer of points and (right) changing the background colour of the reference grid.}
%  \label{fig:ref-adv}
%\end{figure}

@@

Note that the use of local scales needs to be clearly marked. The viewer must realize that the scale is not relative and that big patterns in some locations might be just tiny effects. To indicate this on the locally scaled data plots, one can use an additional aesthetic, like colour, to encode the range of original scale. This is illustrated in Figure~\ref{fig:scaling-col}.

\begin{figure}[htbp]
 \centering
 \includegraphics[width=0.5\linewidth]{month-rescale01-col}%
 \includegraphics[width=0.5\linewidth]{month-rescale01-fill}
 \caption{Adding colour to scaled plots. Here each location has been scaled to range $[0, 1]$ and colour mapped to the range of the original predictions. (Left) Range mapped to colour of line: pale locations have smaller range than darker locations. (Right) Range mapped to fill colour of grid box: blue boxes have smaller ranges. Shapes at all locations are visible, but attention is drawn to locations with larger ranges.}
 \label{fig:scaling-col}
\end{figure}

@@

Climate data is usually large: many locations in space, and many points in time. This makes climate data challenging to work with computationally and to visualize. Plotting the raw data often reveals little more than a well-known observation that ``temperature is quite varied!'' Careful use of models to decompose data into long-term trend, seasonal effects and error is vital for to make the problem both computationally and visually tractable.

@@

  \caption{Glyph-map of raw temperature anomaly data, from 1880-2011. Gaps in the full circle indicate missing values, and thickness of the circle indicates more variability in the measurements.}
  \label{fig:gistemp-raw}
\end{figure}

The raw data shows us missing values and gives some sense of variability, but gives no information about long-term trend. We can look trend by fitting a linear model to each location and displaying the predicted values. Figure~\ref{fig:gistemp-pred} shows this for 1950--2010 (limited time period used due to missing data). Most locations have increasing temperature over this time period. Only a few locations show a decline. These locations tend to be different from their neighbors, which might suggest isolated data problems. The rate of increase differs across regions: steeper inclines in the north and western mountain regions and across the Great Lakes, and flatter inclines in the southeastern USA.

@@

  \caption{Glyph-map of predicted temperature anomaly data, from
    1950-2010. Increasing trends can be seen over most of the USA,
    with varying degrees of increase. Some isolated locations exhibit
    steep declines, and inclines, perhaps indicative of some isolated
    data problems. References line are straight, but look curves because of
    the Z\"ollner optical illusion}
  \label{fig:gistemp-pred}
\end{figure}

@@

\subsection{Irregular locations}
Figure \ref{fig:irregular} shows linear trends for monthly average surface temperature at stations in the USHCN (Version 2) ground station network.  It illustrates three challenges in working with irregular locations: there is no natural choice for icon size, icons can potentially overlap, and comparisons rely more heavily on the reference guides.  The icon sizes were chosen manually for these two plots, although the icons are displayed at about the same size, they correspond to 1$^o$ and 0.5$^o$ square respectively in the geographical coordinates. 
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.55\linewidth]{usa-lin-overlap}%
  \includegraphics[width=0.35\linewidth]{ghcn-mountains}%
  \caption{Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for all continental US USHCN stations (left) and stations in the mountain states (right).} 
  \label{fig:irregular}
\end{figure}
The two panels illustrate one approach to the overlapping problem: zooming.  There is an interplay between icon size (in geographical units), plotting area (in geographical units) and plot size (in display units).  Icons need to be big enough to be readable on the display, but small enough to avoid overlapping.  Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the plotting area (as done in the right panel in Figure \ref{fig:irregular}).  However, there are generally limitations in the maximum display size (or resolution) and the utility of plotting very small areas.

@@

Glyphs can be more general than time series or stars. For multivariate data we may want to represent several variables in the glyphs. Figure~\ref{fig:cloud} shows scatterplots as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled. The bivariate relationship between the two variables is explored spatially, and it is clear that it is different across the region. The equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association. The second plot displays a loess curve fit to the data instead of the individual points, sharpening the view of association. Another option is available in Gauguin \citep{gribov:2006}, a icon based on a histogram, giving small univariate distribution displays for each location.

>>>>

Climate data is composed of measurements, such as temperature, precipitation, and winds, with a spatial and temporal context. The classic display for data of this type of is ``heatmap'', a tiled plot using color to display the value of the variable of interest as each location. When measurements are made at multiple time points it is common to display the data with small multiples, such as in Figure~\ref{fig:nasa-facet}. In this plot, a separate map is drawn for each month (columns) and year (rows), over a six year period of remotely sensed temperature data above Central America \citep{murrell:2010}. Color is used to display de-seasonalized temperature, with red mapped to high values and blue to low. The most noticeable feature is the strong red patch in the equatorial Pacific beginning mid-1997 and tapering out during 1998. This is the El Ni\~no event, a major temperature anomaly. More work is required to see more localized patterns, like the cooler land temperatures seen in early years.

@@

The paper is structured as follows. Section~\ref{sec:construction} describes the algorithm used to create glyphs-maps. Section~\ref{sec:perception} discusses their perceptual properties, including the importance of a visual reference grid, and of carefully consideration of scale. Large data and the interplay of models and data are discussed in Section~\ref{sec:large-data}. Many spatiotemporal data sets have irregular spatial locations, and Section~\ref{sec:irregular} discusses how glyph-maps can be adjusted for this type of data. Three datasets are used for examples:

\begin{quote}
\begin{itemize}

\item[data-expo] The ASA 2009 data expo data \citep{murrell:2010} consists
  of monthly observations of several atmospheric variables from the
  International Satellite Cloud Climatology Project. The dataset
  includes observations over 72 months (1995--2000) on a 24 x 24 grid
  (576 locations) stretching from $113.75^{\circ}$W to
  $56.25^{\circ}$W longitude and $21.25^{\circ}$S to $36.25{^\circ}$N
  latitude.

\item[GISTEMP] surface temperature data provided on $2^{\circ}$ x
  $2^{\circ}$ grid over the entire globe, measured monthly \citep{GISTEMP}. Ground
  station data was de-seasonalized, differenced from from the
  1951-1980 temperature averages, and spatially averaged to obtain
  gridded measurements. For the purposes of this paper, we extracted
  the locations corresponding to the continental USA.

\item[USHCN] (Version 2) ground station network of historical
  temperatures \citep{USHCN}. Temperatures from 1219 stations on the
  contiguous United States, from 1871 to present.

\end{itemize}
\end{quote}

@@

Since the coordinates are linear combinations of two variables, it is possible to build these these line-glyphs interactively, illustrated by software packages that have incorporated tours (as described in \citet{cook:2006}), such as DataViewer \citep{buja:1986}, XGobi \citep{swayne:1991} or GGobi \citep{swayne:2003}. This technique is used to good effect in \citet{buja:1996a}, and was used to examine the climate data in \citet{hobbs:2010}.

@@

An third alternative is to calculate the relevant statistic for each location and display these values as single heatmap. For example, to study long-term trends, we could calculate the slope of a linear model at each location and color tiles across the map by the value of the slope. The disadvantage of this approach is that it is not exploratory, and requires that we have a very precise definition of the measurement we are interested in, and it still suffers from the relative difficultly of color perception.

@@

Each glyph is small, so there is not enough space for a full set of axes. Instead, we add reference lines and boxes. These need to be minimally perceptible, post-attentive \citep{healey} and de-emphasized, so as not to detract from the data. The \textbf{reference grid} is a box, framing each icon, which represents the spatial grid. The \textbf{reference line} is a horizontal line at mid-range. Both help to read differences in slope and intercept. Slope is read by comparing the position of left and right endpoints on the frame, or by reading the angle between the data line and the reference line. Intercept is read from ``average'' position of the line in the box: is it near the top, or near the bottom? Figure~\ref{fig:ref-basic} demonstrates both guides. We draw them in white, so they are minimally perceptible relative to the black data.

@@

%It's also often useful to display other types of reference information. You are only limited by your imagination (and the capabilities of your graphics software), but two simple ideas are to (1) vary the color or transparency of glyphs or references, or (2) use an additional layer to highlight special points. Figure~\ref{fig:ref-adv} illustrates these ideas to display the month with the highest temperature at each locations. Careful color choice is necessary so that this information is perceptible, but not distracting.

%\begin{figure}[htbp]
%  \centering
%  \includegraphics[width=0.5\linewidth]{ref-max-1}%
%  \includegraphics[width=0.5\linewidth]{ref-max-2}
%  \caption{Glyph-maps of seasonal temperature patterns also showing the month with the highest temperature using (left) an additional layer of points and (right) changing the background color of the reference grid.}
%  \label{fig:ref-adv}
%\end{figure}

@@

Note that the use of local scales needs to be clearly marked. The viewer must realize that the scale is not relative and that big patterns in some locations might be just tiny effects. To indicate this on the locally scaled data plots, one can use an additional aesthetic, like color, to encode the range of original scale. This is illustrated in Figure~\ref{fig:scaling-col}.

\begin{figure}[htbp]
 \centering
 \includegraphics[width=0.5\linewidth]{month-rescale01-col}%
 \includegraphics[width=0.5\linewidth]{month-rescale01-fill}
 \caption{Adding intensity and color to scaled plots. Here each location has been scaled to range $[0, 1]$ and color mapped to the range of the original predictions. (Left) Range mapped to intensity of line: pale locations have smaller range than darker locations. (Right) Range mapped to fill color of grid box: blue boxes have smaller ranges. Shapes at all locations are visible, but attention is drawn to locations with larger ranges.}
 \label{fig:scaling-col}
\end{figure}

@@

Climate data is usually large: many locations in space, and many points in time. This makes climate data challenging to work with computationally and to visualize. Careful use of models to decompose data into long-term trend, seasonal effects and error is vital for to make the problem both computationally and visually tractable.

@@

  \caption{Glyph-map of raw temperature anomaly data (GISTEMP), from 1880-2011. Gaps in the full circle indicate missing values, and thickness of the circle indicates more variability in the measurements.}
  \label{fig:gistemp-raw}
\end{figure}

The raw data shows us missing values and gives some sense of variability, but gives no information about long-term trend. We can look at trend by fitting a linear model to each location and displaying the predicted values. Figure~\ref{fig:gistemp-pred} shows this for 1950--2010 (limited time period used due to missing data). Most locations have increasing temperature over this time period. Only a few locations show a decline. These locations tend to be different from their neighbors, which might suggest isolated data problems. The rate of increase differs across regions: steeper inclines in the north and western mountain regions and across the Great Lakes, and flatter inclines in the southeastern USA.

@@

  \caption{Glyph-map of predicted temperature anomaly data (GISTEMP), from
    1950-2010. Increasing trends can be seen over most of the USA,
    with varying degrees of increase. Some isolated locations exhibit
    steep declines, and inclines, perhaps indicative of some isolated
    data problems. References line are straight, but look curves because of
    the Z\"ollner optical illusion}
  \label{fig:gistemp-pred}
\end{figure}

@@

\subsection{Irregular locations}
Figure \ref{fig:irregular} shows linear trends for monthly average surface temperature at stations in the USHCN data.  It illustrates three challenges in working with irregular locations: there is no natural choice for icon size, icons can potentially overlap, and comparisons rely more heavily on the reference guides.  The icon sizes were chosen manually for these two plots, although the icons are displayed at about the same size, they correspond to 1$^o$ and 0.5$^o$ square respectively in the geographical coordinates. 
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.55\linewidth]{usa-lin-overlap}%
  \includegraphics[width=0.35\linewidth]{ghcn-mountains}%
  \caption{Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations (left) and stations in the mountain states (right).} 
  \label{fig:irregular}
\end{figure}
The two panels illustrate one approach to the overlapping problem: zooming.  There is an interplay between icon size (in geographical units), plotting area (in geographical units) and plot size (in display units).  Icons need to be big enough to be readable on the display, but small enough to avoid overlapping.  Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the plotting area (as done in the right panel in Figure \ref{fig:irregular}).  However, there are generally limitations in the maximum display size (or resolution) and the utility of plotting very small areas.

@@

Glyphs can be more general than time series or stars. For multivariate data we may want to represent several variables in the glyphs. Figure~\ref{fig:cloud} shows scatterplots of the data-expo data as icons: temperature values are displayed horizontally and high-cloud values are displayed vertically, both locally scaled. The bivariate relationship between the two variables is explored spatially, and it is clear that it is different across the region. The equatorial Pacific and Caribbean have positive association between temperature and high cloud, while the south American continent has negative association. The second plot displays a loess curve fit to the data instead of the individual points, sharpening the view of association. Another option is available in Gauguin \citep{gribov:2006}, a icon based on a histogram, giving small univariate distribution displays for each location.
